---
title: "Introduction to Health Analytics"
subtitle: "Tutorial 1: Wrangling and visualizing health survey data"
author: ""
date: "Spring 1 2026"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: false
---

# Introduction

This tutorial requires you to use the skills you learned in the first two lectures to manipulate data and plot graphs. We will use data from the **National Health and Nutrition Examination Survey (NHANES)**. This is a large, ongoing survey run by the U.S. Centers for Disease Control and Prevention (CDC). In this tutorial, we will use a simplified version of NHANES data that is bundled in R.

In this session you will practise:

- selecting and filtering variables
- creating new variables 
- making histograms, bar charts, and scatterplots

We introduce a few new functions:

- generating variable labels using conditional logic with `if_else()`
- generating binary variables from continuous numeric variables
- comparing groups using `facet_wrap()`
- adding trend lines using `geom_smooth()`

Load the necessary packages.
```{r, include=FALSE}
packages = (c("tidyverse", "NHANES"))

## Now load or install&load all
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE, repos = "https://cloud.r-project.org")
      library(x, character.only = TRUE)
    }
  }
)

```

Start by loading the NHANES data into a new object called `df`.
```{r}
df <- NHANES
head(df)
```

We will work with the following variables:

- **Age**: age in years.
- **Gender**: `"male"` or `"female"`.
- **BPSysAve**: *average systolic blood pressure* (mm Hg) and  **BPDiaAve**: *average diastolic blood pressure* (mm Hg).  Blood pressure is usually written as two numbers, like **120/80**.  The **systolic** pressure (first number) measures pressure in the arteries when the heart **contracts** and the **diastolic** pressure (second number) measures pressure when it relaxes. High blood pressure is a major risk factor for heart disease and stroke.
- **BMI** (*Body Mass Index*):  BMI is calculated as  $\text{BMI} = \frac{\text{weight (kg)}}{\text{height (m)}^2}$ It is commonly used as a (very) rough indicator of healthy weight in population health studies.
- **BMI_WHO**: BMI grouped into categories based on World Health Organization cut-offs  
  (e.g. underweight, normal weight, overweight, obese).
- **Diabetes**: indicator of whether a participant has diabetes.


# Question 1: Create a clean analysis dataset
1. Before plotting, we will keep only the variables we care about and remove observations with missing values. Create a dataset called `df_small` that:

- Keeps only  
   `ID, Age, Gender, BPSysAve, BMI, BMI_WHO, Diabetes`
- Removes rows where BMI, systolic blood pressure, or gender are missing.

2. How many observations were filtered out due to missing data?

```{r}

```

# Question 2: Creating binary variables
Blood pressure is a continuous variable, but people are often categorized as having or not having **hypertension** (high blood pressure) based on a cut-off value. In this question we will create a **binary variable** indicating hypertension. For this exercise we define hypertension as systolic blood pressure greater than or equal to 130 mm Hg.

We can use`if_else()` to create a new variable based on a condition. The general structure is:

```
if_else(condition, value_if_true, value_if_false)
```

1. Create a binary variable `Hypertension`. This should be equal to 1 if `BPSysAve` is greater than or equal to 130, and 0 otherwise. 
2. What percentage of people in the sample have hypertension?

```{r}

```

# Question 3: Histograms
Hisograms are useful for visualizing the distribution of a continuous variable.

1. Plot a histogram of systolic blood pressure. Show males and females in **separate panels**. HINT: use `facet_wrap()` to splits one plot into multiple panels based on a grouping variable. 
2. CHALLENGE: Add a line to show the cutoff threshold for hypertension

```{r}

```

# Question 4: Bar charts 

Bar charts are useful for counts and group comparisons.

1. Make a bar chart showing number of people in each of the `BMI_WHO` categories. Exclude the people who have missing BMI category.
2. Make a stacked bar chart showing number of people with and without hypertension in each BMI category. HINT: use the `fill` aesthetic of geom_bar to fill bars based on hypertension status. You can only fill by categorical variables, so you will need to covert `Hypertension` to a factor variable i.e. `fill=factor(Hypertension)`.
3. CHALLENGE: Make a bar chart showing percent of people in each BMI category with and without hypertension. Note that you should be able to do this by setting `position="fill"` in the code for the stacked bar chart you just made.

```{r}

```

```{r}

```

```{r}

```
  
# Question 5: Scatterplots by sub-group

A scatterplot helps us see whether two continuous variables are related.

1. Make a scatterplot showing the relationship between BMI (x-axis) and systolic blood pressure (y-axis). Show all points in the same panel, but plot males and females with different colors.
2. Make plots showing the relationship between BMI (x-axis) and systolic blood pressure (y-axis) for males and females separately using `facet_wrap()`.

```{r}

```

```{r}

```

# Question 6: Adding trend lines 
In this question we will explore using `geom_smooth()` to add a fitted line to a plot to show the average relationship between two variables, using either a default smoother or a model you specify (e.g. a linear regression).

1. Add a linear fit to the scatterplot of BMI vs systolic blood pressure for all participants by adding a layer `geom_smooth(method = "lm")` to the scatter plot you made in question 5.1 above. You will see this fits separate lines for males and females.
2. Instead of fitting separate lines for males and females, set the same line for everyone. HINT: specify the mapping of Gender to color **inside** the `geom_point()` function, not in the main `ggplot()` function.
2. What happens if you don't specify `method = "lm"`? Try it out by adding a default smoother to the same scatterplot.

```{r}

```

```{r}

```
```{r}

```


# Question 7: Do some exploratory plotting!

Create **one clear and informative figure** that shows something interesting about:

- diabetes and BMI
- diabetes and blood pressure
- or how diabetes, BMI, and gender interact
 
 

