---
title: "Introduction to Health Analytics"
subtitle: "Linear regression 1"
author: ""
date: "Spring 1 2026"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
---

# Introduction

This exercise will guide you through running linear regression in R and interpreting the results. We will use the `NHANES` dataset available from the `R` package `NHANES`. We will study the relationship between **education** and **mental health**.

- $y$ is `DaysMentalHlthBad` is the number of days in the past 30 days when respondent \(i\) reported their mental health was “not good”.  (In NHANES this is often coded as a 0–30 count with many zeros.)
- $x$ is is `Education`, coded as categories.
- You will be using gender and age as covariates

As always start by loading the necessary packages. We're going to use `tidyverse` for data manipulation and visualization, `NHANES` to load the NHANES data, and `modelsummary` and `kableExtra` and `knitr` to produce nice looking tables.

```{r, include=FALSE, results='hide'}
packages <- c("tidyverse", "broom", "tidyr","srvyr", "NHANES", "modelsummary", "kableExtra")
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)
```

Load the NHANES data into an R dataframe called `nhanes_data`.

```{r}
# View structure of the dataset
nhanes_data <- NHANES
```

# Data cleaning and summary statistics

1. Filter the data to exclude people who have `NA` recorded for any of the following variables:  `Education`, `DaysMentHlthBad`, `Gender` or `Age`.  
2. Create two new binary variables:

  - `CollegeGrad` equal to 1 if `Education` is "College Grad" and zero otherwise. 
  - `Male` equal to 1 if `Gender` is "male" and zero otherwise. 

3. For the filtered observations calculate:

  a) What percent of people have `DaysMentHlthBad` equal to zero? 
  b) What is the mean and standard deviation of `DaysMentHlthBad` across all observations?  
  c) What is the mean and standard deviation of `DaysMentHlthBad` for people with `CollegeGrad` equal to 1 and for people with `CollegeGrad` equal to 0?

```{r wrangle}
```


```{r descriptives}
```

# Data visualization
Using the filtered data from question 1, generate a bar chart showing the average number of days of bad mental health (`DaysMentHlthBad`) by the new `CollegeGrad' variable. Plot density bars side by side for college graduates and non-college graduates. Would a log transformation of \(Y\) make sense here? Why or why not?

```{r barplot}
```

# Univariate regression

1. Estimate a linear regression where `DaysMentHlthBad` is the dependent variable and the explanatory variable is the `CollegeGrad` dummy variable. Use `summary()` to show the coefficients and from your estimated model.

2. How do you interpret the coefficient on CollegeGrad? In your answer, comment on the direction of the relationship and the magnitude of the effect? Ignore statistical uncertainty for now.

3. Explain why the coefficient on `CollegeGrad` in the model might not be a causal estimate of the effect of a college education on mental health. 

```{r linear}
```

# Multivariate regression

1. Estimate a linear regression where `DaysMentHlthBad` is the dependent variable, and the explanatory variables are `CollegeGrad`, `Age`, and `Male`. 

2. How do you interpret the coefficients on `CollegeGrad`, `Age` and `Male`? Ignore statistical uncertainty for now.

3. Predict the number of days of bad mental health for (i) a 35-year old female with a college degree and (ii) a 35-year old male with a college degree.

```{r linearmv}
```
# Interactions

1. Estimate a linear regression where `DaysMentHlthBad` is the dependent variable, and the explanatory variables are `CollegeGrad`, `Age`, `Male`. and an interaction between `CollegeGrad` and `Male`.How do you interpret the coefficients on `CollegeGrad`, `Age` and `Male`?

2. How do you interpret the coefficients on `CollegeGrad`, `Male` and `CollegeGrad`\(\times\)`Male`? Ignore statistical uncertainty for now.

3. Predict the number of days of bad mental health for (i) a 35-year old female with a college degree and (ii) a 35-year old male with a college degree.

```{r linearmvint}

```

# Regression tables

1. You can put the results from all three regressions side-by-side in a regression table as follows. NOTE: you can get this to run by deleting `eval=FALSE` once you have estimated your models. You may need to change the list of models to match your model object names.

```{r table, eval=FALSE}
modelsummary(
  list("Model 1" = lm1, "Model 2" = lm2, "Model 3" = lm3),
  output = "kableExtra",
  statistic = "({std.error})",   # show SEs in parentheses
  stars = TRUE
)
```