---
title: "Introduction to Health Analytics"
subtitle: "Linear regression 2: Uncertainty and hypothesis testing"
author: "Sam Burn"
date: "Spring 1 2026"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

# Introduction

This exercise builds directly on the lecture 5 exercise. You will run the same regressions of mental health on education (and other covariates) using the `NHANES` data, but now the focus is on **uncertainty and hypothesis testing**:

- Read off and interpret t-statistics, p-values, and F-statistics from regression output.
- **Manually** compute t-statistics and p-values for a couple of null hypotheses, including a non-zero null.
- Use `linearHypothesis()` to test joint hypotheses and linear combinations of coefficients (this last one becomes especially useful with interaction terms).

Outcome and regressors are the same as in the lecture 5 exercise:

- $y$ is `DaysMentHlthBad`, the number of days in the past 30 days when respondent $i$ reported their mental health was "not good".
- $x$ is `CollegeGrad`, a binary indicator for whether the respondent's `Education` is "College Grad".
- Covariates: `Age` (years) and `Male` (binary).

```{r, include=FALSE}
packages <- c("tidyverse", "broom", "NHANES", "modelsummary", "knitr", "kableExtra", "car")
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)
```

# Data setup

The code below loads NHANES, filters out missing values, and creates the `CollegeGrad` and `Male` indicators. We don't repeat the EDA / wrangling questions from the lecture 5 exercise — assume that's already done.

```{r data-setup}
nhanes_data <- NHANES

nhanes_clean <- nhanes_data %>%
  filter(!is.na(Education),
         !is.na(Age),
         !is.na(DaysMentHlthBad),
         !is.na(Gender)) %>%
  mutate(
    CollegeGrad = ifelse(Education == "College Grad", 1, 0),
    Male        = ifelse(Gender == "male", 1, 0)
  )
```

# Univariate regression and hypothesis testing

1. Estimate a linear regression with `DaysMentHlthBad` as the dependent variable and `CollegeGrad` as the only explanatory variable. Use `summary()` to inspect the output.
2. Read off the t-statistic and p-value for `CollegeGrad` directly from the regression output. What null hypothesis are these testing? Is the test one-sided or two-sided? At the 5% level, can we reject the null?
3. **Manually compute** the t-statistic for the null hypothesis $H_0: \beta_{\text{CollegeGrad}} = 0$ from the point estimate and standard error. Confirm it matches what R reports.
4. **Manually compute** the two-sided p-value using `pt()` (Student-t distribution with $n - k - 1$ degrees of freedom). Confirm it matches.

```{r univariate-test}

```

# Hypothesis test with a non-zero null

Earlier work suggests college graduates have somewhat fewer bad mental health days. Suppose a colleague claims the "true" effect is exactly $-2$ days. Test the null hypothesis $H_0: \beta_{\text{CollegeGrad}} = -2$ against the two-sided alternative.

1. **Manually compute** the t-statistic for $H_0: \beta_{\text{CollegeGrad}} = -2$.
2. **Manually compute** the corresponding two-sided p-value.
3. At the 5% level, do you reject the null?
4. Confirm your answer using `linearHypothesis(lm1, "CollegeGrad = -2")` from the `car` package. The reported F-statistic should equal the square of your manual t-statistic.

```{r non-zero-null}

```

# Multivariate regression and the F-test

1. Estimate the multivariate regression of `DaysMentHlthBad` on `CollegeGrad`, `Age`, and `Male`.
2. The F-statistic at the bottom of `summary()` tests $H_0: \beta_{\text{CollegeGrad}} = \beta_{\text{Age}} = \beta_{\text{Male}} = 0$ jointly. Read it off and decide whether we reject.
3. Use `linearHypothesis()` to test only $H_0: \beta_{\text{Age}} = \beta_{\text{Male}} = 0$ (i.e. the demographic controls jointly add nothing once we have `CollegeGrad`).

```{r multivariate-ftest}

```

# Interactions and linear combinations

The lecture 5 exercise added an interaction between `CollegeGrad` and `Male`. The interesting hypothesis tests now require **linear combinations** of coefficients.

1. Estimate the regression with `CollegeGrad`, `Age`, `Male`, and the interaction `CollegeGrad:Male`.
2. The coefficient on `CollegeGrad` represents the effect of being a college graduate **for women** (since `Male = 0` for women). Test $H_0: \beta_{\text{CollegeGrad}} = 0$.
3. The effect of being a college grad **for men** is $\beta_{\text{CollegeGrad}} + \beta_{\text{CollegeGrad:Male}}$. Test $H_0: \beta_{\text{CollegeGrad}} + \beta_{\text{CollegeGrad:Male}} = 0$ using `linearHypothesis()`.
4. Test whether the effect of college differs between men and women i.e. $H_0: \beta_{\text{CollegeGrad:Male}} = 0$.

```{r interactions}

```

# Confidence intervals

1. Compute a 95% confidence interval for $\beta_{\text{CollegeGrad}}$ from `lm3` **manually**: $\hat{\beta} \pm t_{\alpha/2, n - k - 1} \cdot \mathrm{SE}(\hat{\beta})$, using the t-critical value.
2. Compare to `confint(lm3)`.
3. Use the manual CI to confirm the test of $H_0: \beta_{\text{CollegeGrad}} = -2$ (does $-2$ fall inside the 95% CI?).

```{r confidence-intervals}

```
