---
title: "Introduction to Health Analytics"
subtitle: "Linear regression 2: Uncertainty and hypothesis testing"
author: "Sam Burn"
date: "Spring 1 2026"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

# Introduction

This exercise builds directly on the lecture 5 exercise. You will run the same regressions of mental health on education (and other covariates) using the `NHANES` data, but now the focus is on **uncertainty and hypothesis testing**:

- Read off and interpret t-statistics, p-values, and F-statistics from regression output.
- **Manually** compute t-statistics and p-values for a couple of null hypotheses, including a non-zero null.
- Use `linearHypothesis()` to test joint hypotheses and linear combinations of coefficients (this last one becomes especially useful with interaction terms).

Outcome and regressors are the same as in the lecture 5 exercise:

- $y$ is `DaysMentHlthBad`, the number of days in the past 30 days when respondent $i$ reported their mental health was "not good".
- $x$ is `CollegeGrad`, a binary indicator for whether the respondent's `Education` is "College Grad".
- Covariates: `Age` (years) and `Male` (binary).

```{r, include=FALSE}
packages <- c("tidyverse", "broom", "NHANES", "modelsummary", "knitr", "kableExtra", "car")
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)
```

# Data setup

The code below loads NHANES, filters out missing values, and creates the `CollegeGrad` and `Male` indicators. We don't repeat the EDA / wrangling questions from the lecture 5 exercise — assume that's already done.

```{r data-setup}
nhanes_data <- NHANES

nhanes_clean <- nhanes_data %>%
  filter(!is.na(Education),
         !is.na(Age),
         !is.na(DaysMentHlthBad),
         !is.na(Gender)) %>%
  mutate(
    CollegeGrad = ifelse(Education == "College Grad", 1, 0),
    Male        = ifelse(Gender == "male", 1, 0)
  )
```

# Univariate regression and hypothesis testing

1. Estimate a linear regression with `DaysMentHlthBad` as the dependent variable and `CollegeGrad` as the only explanatory variable. Use `summary()` to inspect the output.
2. Read off the t-statistic and p-value for `CollegeGrad` directly from the regression output. What null hypothesis are these testing? Is the test one-sided or two-sided? At the 5% level, can we reject the null?
3. **Manually compute** the t-statistic for the null hypothesis $H_0: \beta_{\text{CollegeGrad}} = 0$ from the point estimate and standard error. Confirm it matches what R reports.
4. **Manually compute** the two-sided p-value using `pt()` (Student-t distribution with $n - k - 1$ degrees of freedom). Confirm it matches.

```{r univariate-test}
lm1 <- lm(DaysMentHlthBad ~ CollegeGrad, data = nhanes_clean)
summary(lm1)

# Pull out the estimate and SE for CollegeGrad
b_cg <- coef(lm1)["CollegeGrad"]
se_cg <- summary(lm1)$coefficients["CollegeGrad", "Std. Error"]
n     <- nobs(lm1)
k     <- length(coef(lm1)) - 1  # number of regressors (excluding intercept)

# Manual t-statistic for H0: beta = 0
t_zero <- (b_cg - 0) / se_cg

# Two-sided p-value using the t-distribution with n - k - 1 d.f.
p_zero <- 2 * pt(abs(t_zero), df = n - k - 1, lower.tail = FALSE)

c(estimate = b_cg, se = se_cg, t = t_zero, p = p_zero)
```

**Answer:** Both the t-statistic ($\approx -7.1$) and the (essentially zero) p-value match what R reports in the `summary()` table. R's default test is two-sided and tests $H_0: \beta_{\text{CollegeGrad}} = 0$. We reject the null at the 5% level.

# Hypothesis test with a non-zero null

Earlier work suggests college graduates have somewhat fewer bad mental health days. Suppose a colleague claims the "true" effect is exactly $-2$ days. Test the null hypothesis $H_0: \beta_{\text{CollegeGrad}} = -2$ against the two-sided alternative.

1. **Manually compute** the t-statistic for $H_0: \beta_{\text{CollegeGrad}} = -2$.
2. **Manually compute** the corresponding two-sided p-value.
3. At the 5% level, do you reject the null?
4. Confirm your answer using `linearHypothesis(lm1, "CollegeGrad = -2")` from the `car` package. The reported F-statistic should equal the square of your manual t-statistic.

```{r non-zero-null}
# Manual t-statistic for H0: beta = -2
t_neg2 <- (b_cg - (-2)) / se_cg
p_neg2 <- 2 * pt(abs(t_neg2), df = n - k - 1, lower.tail = FALSE)

c(t = t_neg2, p = p_neg2)

# Confirm with linearHypothesis
linearHypothesis(lm1, "CollegeGrad = -2")
```

**Answer:** The point estimate is around $-1.55$, only $\approx 0.45$ above $-2$, with SE $\approx 0.2$. The manual t-statistic is around $2$ and the two-sided p-value is roughly $0.04$, so we cannot reject the colleague's claim at the 5% level. `linearHypothesis()` reports $F \approx 4$, which equals $t^2$ so this is equivalent!

# Multivariate regression and the F-test

1. Estimate the multivariate regression of `DaysMentHlthBad` on `CollegeGrad`, `Age`, and `Male`.
2. The F-statistic at the bottom of `summary()` tests $H_0: \beta_{\text{CollegeGrad}} = \beta_{\text{Age}} = \beta_{\text{Male}} = 0$ jointly. Read it off and decide whether we reject.
3. Use `linearHypothesis()` to test only $H_0: \beta_{\text{Age}} = \beta_{\text{Male}} = 0$ (i.e. the demographic controls jointly add nothing once we have `CollegeGrad`).

```{r multivariate-ftest}
lm2 <- lm(DaysMentHlthBad ~ CollegeGrad + Age + Male, data = nhanes_clean)
summary(lm2)

# Joint test of Age = Male = 0
linearHypothesis(lm2, c("Age = 0", "Male = 0"))
```

**Answer:** The overall F-statistic at the bottom of `summary(lm2)` is around 45 with $p < 2 \times 10^{-16}$, so we strongly reject the joint null that all three coefficients are zero. The restricted F-test of $\beta_{\text{Age}} = \beta_{\text{Male}} = 0$ also gives a large F-statistic with a tiny p-value.

# Interactions and linear combinations

The lecture 5 exercise added an interaction between `CollegeGrad` and `Male`. The interesting hypothesis tests now require **linear combinations** of coefficients.

1. Estimate the regression with `CollegeGrad`, `Age`, `Male`, and the interaction `CollegeGrad:Male`.
2. The coefficient on `CollegeGrad` represents the effect of being a college graduate **for women** (since `Male = 0` for women). Test $H_0: \beta_{\text{CollegeGrad}} = 0$. 
3. The effect of being a college grad **for men** is $\beta_{\text{CollegeGrad}} + \beta_{\text{CollegeGrad:Male}}$. Test $H_0: \beta_{\text{CollegeGrad}} + \beta_{\text{CollegeGrad:Male}} = 0$ using `linearHypothesis()`. 
4. Test whether the effect of college differs between men and women i.e. $H_0: \beta_{\text{CollegeGrad:Male}} = 0$. 

```{r interactions}
lm3 <- lm(DaysMentHlthBad ~ CollegeGrad * Male + Age, data = nhanes_clean)
summary(lm3)

# Effect of college for women: H0: CollegeGrad = 0
linearHypothesis(lm3, "CollegeGrad = 0")

# Effect of college for men: H0: CollegeGrad + CollegeGrad:Male = 0
linearHypothesis(lm3, "CollegeGrad + CollegeGrad:Male = 0")
```

**Answer:**

- For women, we can reject the null hypothesis that $H_0: \beta_{\text{CollegeGrad}} = 0$ at the 0.1% significance level: college grad women have on average around 1.76 fewer bad mental health days, $p < 0.001$.
- For men, the linear combination is around $-1.45$ ($-1.76 + 0.32$). We can reject the null hypothesis that $H_0: \beta_{\text{CollegeGrad}} + \beta_{\text{CollegeGrad:Male}} = 0$ at the 0.1% significance level.
- The interaction term itself has $p \approx 0.47$. We cannot reject that the effect of college on mental health is the same for men and women. So while college matters for both groups individually, the data don't support the claim that it matters *more* for one group than the other.

# Confidence intervals

1. Compute a 95% confidence interval for $\beta_{\text{CollegeGrad}}$ from `lm3` **manually**: $\hat{\beta} \pm t_{\alpha/2, n - k - 1} \cdot \mathrm{SE}(\hat{\beta})$, using the t-critical value.
2. Compare to `confint(lm3)`.
3. Use the manual CI to confirm the test of $H_0: \beta_{\text{CollegeGrad}} = -2$ (does $-2$ fall inside the 95% CI?).

```{r confidence-intervals}
b_cg3  <- coef(lm3)["CollegeGrad"]
se_cg3 <- summary(lm3)$coefficients["CollegeGrad", "Std. Error"]
df_lm3 <- nobs(lm3) - length(coef(lm3))
tcrit  <- qt(0.975, df = df_lm3)

ci_manual <- c(b_cg3 - tcrit * se_cg3, b_cg3 + tcrit * se_cg3)
ci_manual

confint(lm3, "CollegeGrad", level = 0.95)
```

**Answer:** Manual and `confint()` agree to several decimal places. The interval roughly spans $-2.2$ to $-1.3$, so $-2$ is outside the CI — consistent with the earlier rejection of $H_0: \beta = -2$.

